Supplementary material

### 2018-08-15 paper/discussion.tex
missing the conclusions

Removing the geographical normalization/interpolation step greatly reduces the
amount of work necessary, but the method still requires consistent variables
across the years. Matching the fields of the public census can be trivial for
some aspects, like Age, but challenging for others, like Income. The divulged
income ranges vary over time, the actual value changes due to inflation and
other factors, and so on. Moreover, some fields were not considered in earlier
censuses, such as Race in Canada, or Hispanic population in the USA, hampering its
use when they are available. We matched some aspects, but a deeper demographic
analysis would greatly benefit from all available information.

>>>>

Removing the geographical normalization/interpolation step greatly reduces the
amount of work necessary, but the method still requires consistent variables
across the years. Matching the fields of the public census can be trivial for
some aspects (Age), but challenging for others (Income). The divulged income
ranges vary over time, the actual value changes due to inflation and other
factors, and so on. Moreover, some fields were not considered in earlier
censuses, such as Race in Canada, or Hispanic population in the USA, hampering
its use when they are available. Since this is only a prototype, we matched few
aspects, but a deeper demographic analysis would greatly benefit from all
available information.

The limitation on the number of displayed clusters because of the limited number
of distinguishable colors was significant. While increasing the number of
clusters would further complicate an already complex analysis, it might be
warranted for some regions. Color is a fundamental and intuitive tool for
information representation that can be coherently used across different plots,
so we opted to use it, even if in a limited way. With eight colors, there was
overlap between some clusters, the relevance gradient, and the color
combination.

### 2018-08-15 paper/intro.tex
one past until LA
missing the conclusions

These new datasets usually contain GPS coordinates for the records, leading to
\emph{point-based} data. Combined with the corresponding timestamps, this data
is easily suitable for longitudinal analysis. But most demographic datasets are
\emph{region-based}, where the measurements are associated with pre-defined
regions, not only for an additional level of privacy protection, but because
some measurements only make sense over a defined area. Census data is a classic
example of this format, with datasets available from 1790 onwards for the
US~\cite{nhgis}. Despite this unmatched temporal availability, longitudinal
analyses of census data are often restricted in time, especially when smaller
tabulation areas are considered, such as census tracts (CT) or dissemination
areas, which evolve to reflect changes in population density, leading to
geographic inconsistencies across time, and the traditional time-series based
approach is no longer viable.

However, these analyses are necessary to understand the urban environment.
Indeed, two different regions can have similar average income for a given year,
while one is experiencing a process of economic improvement and the other one
impoverishment. Quite obviously, a single snapshot cannot be used to identify
gentrification, migration, education changes, or any of the relevant processes
that happen over time.

@@

The main contribution of this application paper is an visualization-based
alternative to the geographical harmonization, a combination of established
graph based processing and information visualization techniques allowing
tract-by-tract comparison, the identification and visualization of patterns of
demographic evolution without geographic harmonization, effectively removing one
of the most challenging problems in longitudinal demographic analysis. We also
include illustrative scenarios and our prototype is available at
\url{http://uoft.me/piccard}, including more than fourty regions in the US and
Canada.

% We adopt census data as familiar operational
% context, but the proposed methodology is suitable for most region based
% measurements.

>>>>

These new datasets usually contain GPS coordinates for the records, leading to
\emph{point-based} data. Combined with the corresponding timestamps, this data
is easily suitable for longitudinal analysis. But most demographic datasets are
\emph{region-based}, where the measurements are associated with pre-defined
regions, not only for an additional level of privacy protection, but because
some measurements only make sense over a defined area. Census data is a classic
example of this format, with datasets available from 1790 onward for the
US~\cite{nhgis}. Despite this unmatched temporal availability, longitudinal
analyses of census data are often restricted in time, especially when smaller
tabulation areas are considered, such as census tracts (CT) or dissemination
areas, which evolve to reflect changes in population density, leading to
geographic inconsistencies across time, and the traditional time-series based
approach is no longer viable.

However, these analyses are necessary to understand the urban environment.
Indeed, two different regions can have similar average income for a given year,
while one is experiencing a process of economic improvement and the other one
impoverishment. A single snapshot cannot be used to identify gentrification,
migration, education changes, or any of the relevant processes that happen over
time.

@@

The main contribution of this application paper is an visualization-based
alternative to the geographical harmonization, a combination of established
graph based processing and information visualization techniques allowing
tract-by-tract comparison, the identification and visualization of patterns of
demographic evolution without geographic harmonization, effectively removing one
of the most challenging problems in longitudinal demographic analysis. We also
include illustrative scenarios and our prototype is available at
\url{http://uoft.me/piccard}, including more than forty regions in the US and
Canada. The source code is publicly available at
\url{https://github.com/fabioasdias/piccard}.

### 2018-08-15 paper/method.tex
missing the conclusions

Each resulting cluster is contiguous in the graph. This means that two similar,
but non-contiguous, sets of CTs will be classified into two different clusters,
which can be counterintuitive. To overcome this issue, we \emph{augment} the
graph with two new edges per node from a nearest neighbors
graph~\cite{scikit-learn} using only the distances between the histograms. These
edges connect nodes with similar content, if they are not already connected,
providing a path for the algorithm to group similar nodes. Theoretically, adding
more of these content based edges could be used to decrease the impact of the
spatio-temporal edges, controling the balance between content and topology in
the result. In practice, the effect is dependent on the data itself, and the
results are not consistent, or predictable, across different cities. We fixed it
at two edges because it was the lowest number that empircally led to consistent
clusters, but we believe that this idea warrants further investigation, as an
alternative to mixing parameters in the distance metric~\cite{Chavent2017}.

@@

The bottom part of the interface contains the details for the selected
trajectories, or for the whole city if nothing is selected, as illustrated in
Figure~\ref{fig:ui}. This panel contains two main regions: the small multiple
maps, depicting the clusters at each year, and the stacked bar plots that
summarize the overall composition of these regions. Some finer localization
information is lost using small multiples, such as small border changes, but
that information is available at the larger map. All the maps are linked with
syncronized navigation, and the use of small multiples allows the exploration of
each temporal census individually, and its comparison to the others, with
minimal interaction.

>>>>

Each resulting cluster is contiguous in the graph. This means that two similar,
but non-contiguous, sets of CTs will be classified into two different clusters,
which can be counter-intuitive. To overcome this issue, we \emph{augment} the
graph with two new edges per node from a nearest neighbors
graph~\cite{scikit-learn} using only the distances between the histograms. These
edges connect nodes with similar content, if they are not already connected,
providing a path for the algorithm to group similar nodes. Theoretically, adding
more of these content based edges could be used to decrease the impact of the
spatio-temporal edges, controlling the balance between content and topology in
the result. In practice, the effect is dependent on the data itself, and the
results are not consistent, or predictable, across different cities. We fixed it
at two edges because it was the lowest number that empirically led to consistent
clusters, but we believe that this idea warrants further investigation, as an
alternative to mixing parameters in the distance metric~\cite{Chavent2017}.

@@

The bottom part of the interface contains the details for the selected
trajectories, or for the whole city if nothing is selected, as illustrated in
Figure~\ref{fig:ui}. This panel contains two main regions: the small multiple
maps, depicting the clusters at each year, and the stacked bar plots that
summarize the overall composition of these regions. Some finer localization
information is lost using small multiples, such as small border changes, but
that information is available at the larger map. All the maps are linked with
synchronized navigation, and the use of small multiples allows the exploration of
each temporal census individually, and its comparison to the others, with
minimal interaction.

### 2018-08-15 paper/related.tex
missing the conclusions

Since our problem emcompasses several fields, we divided this section into
specific subproblems: \emph{longitudinal demographic studies}, describing the
traditional approach to perform longitudinal studies; \emph{data
representation}, exploring how evolving geographic data can be represented for
processing; \emph{Data clustering}, briefly reviewing existing clustering
methods; and \emph{cluster characterization}, exploring how the clusters can be
visually summarized.

\subsection{Longitudinal demographic studies}
Census data is used not only to discover demographic
patterns~\cite{Firebaugh2016}, but to correlate demographic charateristics to
other measurements~\cite{diez1997neighborhood}. However, longitudinal studies
are rare: \emph{"(...) One of the most challenging and fascinating areas in
spatial statistics is the synthesis of spatial data collected at different
spatial scales(...)"}~\cite{gotway2002combining}.

While CT level data is readily available for the US since 1910~\cite{nhgis},
most studies consider the period between 1970 and 2010, using pre-harmonized
data~\cite{Logan2014,nhgis}. Despite the inherent
errors~\cite{Logan2016,Hallisey2017}, this dataset became the standard source
for longitudinal demographic data, with similar efforts appearing in other
countries~\cite{Liu2015,Lee2015,Allen2018}. This result was significant for the
field, but it also restricts the useable data, since new datasets need to be
similarly processed.

Another option considers the use of grid data~\cite{Dmowska2017,Dmowska2018},
where small rectangular areas are used, in an approach similar to satellite
imagery. Beyond the increased spatial accuracy, this approach does not require
complex harmonizations when new data is considered. However, demographic data is
usually not available in this format, especially from older sources, and the
conversion from tabulation areas can introduce significant errors.

@@

Indeed, one crucial step in most clustering algorithms is the definition of the
number of clusters. We sidestep this problem by considering hierarchical
methods~\cite{soille2012morphological}, where the result is not a partition of
the data, but a tree of partitions. This approach is interesting for interactive
methods, because it allows the user to change the number of displayed clusters
with minimal processing. Since our data is represented as a graph, one option
would be the watershed cuts algorithm~\cite{Cousty2009}, inspired by the well
known image processing segmentation and equally prone to oversegmentation.
Considering that the processing time is also a relevant factor, we opted for an
heuristic variation of the maximum weighted matching algorithm called
\emph{sorted maximal matching}~\cite{markus2017}, which merges clusters based on
the weights of the edges between pairs of clusters.

\subsection{Cluster characterization}
Visually representing evolving spatial data is a challenging old
problem~\cite{monmonier1990strategies,andrienko2003exploratory,ferreira2015visual,Zheng2016}.
Most geographic data is naturally bidimensional and maps work well in this
case~\cite{Zheng2016,ward2015interactive}, but the temporal dimension cannot be
so naturally represented. One straightforward option is to leverage
tridimensional plots~\cite{andrienko2014visualization,Tominski2012a}, but this
can lead to visual obstructions or scaling problems unless a tridimensional
display device is used. Animation can also be explored in some specific
cases~\cite{buschmann2014real}, but it is not a general approach. Glyphs can
also be used~\cite{seebacher2017visual,Andrienko2017}, but this may lead to
cluttering when many small regions are present. A simpler, well adopted, option
is to display a map that corresponds to a subset of the temporal information,
allowing the user to change the time with an associated
control~\cite{Chen2017,Valdivia2015,Alce2018,Doraiswamy2014}. Small multiples
can be used~\cite{VonLandesberger2016}, but only when there are few temporal
snapshots. However, none of these options is suitable to represent many
variables at the same time.

>>>>

Since our problem encompasses several fields, we divided this section into
specific sub problems: \emph{longitudinal demographic studies}, describing the
traditional approach to perform longitudinal studies; \emph{data
representation}, exploring how evolving geographic data can be represented for
processing; \emph{Data clustering}, briefly reviewing existing clustering
methods; and \emph{cluster characterization}, exploring how the clusters can be
visually summarized.

\subsection{Longitudinal demographic studies}
Census data is used not only to discover demographic
patterns~\cite{Firebaugh2016}, but to correlate demographic characteristics to
other measurements~\cite{diez1997neighborhood}. However, longitudinal studies
are rare: \emph{"(...) One of the most challenging and fascinating areas in
spatial statistics is the synthesis of spatial data collected at different
spatial scales(...)"}~\cite{gotway2002combining}.

While CT level data is readily available for the US since 1910~\cite{nhgis},
most studies consider the period between 1970 and 2010, using pre-harmonized
data~\cite{Logan2014,nhgis}. Despite the inherent
errors~\cite{Logan2016,Hallisey2017}, this dataset became the standard source
for longitudinal demographic data, with similar efforts appearing in other
countries~\cite{Liu2015,Lee2015,Allen2018}. This result was significant for the
field, but it also restricts the usable data, since new datasets need to be
similarly processed.

Another option considers the use of grid data~\cite{Dmowska2017,Dmowska2018},
where small rectangular areas are used, in an approach similar to satellite
imagery. Beyond the increased spatial accuracy, this approach does not require
complex harmonization when new data is considered. However, demographic data is
usually not available in this format, especially from older sources, and the
conversion from tabulation areas can introduce significant errors.

@@

Indeed, one crucial step in most clustering algorithms is the definition of the
number of clusters. We sidestep this problem by considering hierarchical
methods~\cite{soille2012morphological}, where the result is not a partition of
the data, but a tree of partitions. This approach is interesting for interactive
methods, because it allows the user to change the number of displayed clusters
with minimal processing. Since our data is represented as a graph, one option
would be the watershed cuts algorithm~\cite{Cousty2009}, inspired by the well
known image processing segmentation and equally prone to over segmentation.
Considering that the processing time is also a relevant factor, we opted for an
heuristic variation of the maximum weighted matching algorithm called
\emph{sorted maximal matching}~\cite{markus2017}, which merges clusters based on
the weights of the edges between pairs of clusters.

\subsection{Cluster characterization}
Visually representing evolving spatial data is a challenging old
problem~\cite{monmonier1990strategies,andrienko2003exploratory,ferreira2015visual,Zheng2016}.
Most geographic data is naturally bi dimensional and maps work well in this
case~\cite{Zheng2016,ward2015interactive}, but the temporal dimension cannot be
so naturally represented. One straightforward option is to leverage
tridimensional plots~\cite{andrienko2014visualization,Tominski2012a}, but this
can lead to visual obstructions or scaling problems unless a tridimensional
display device is used. Animation can also be explored in some specific
cases~\cite{buschmann2014real}, but it is not a general approach. Glyphs can
also be used~\cite{seebacher2017visual,Andrienko2017}, but this may lead to
cluttering when many small regions are present. A simpler, well adopted, option
is to display a map that corresponds to a subset of the temporal information,
allowing the user to change the time with an associated
control~\cite{Chen2017,Valdivia2015,Alce2018,Doraiswamy2014}. Small multiples
can be used~\cite{VonLandesberger2016}, but only when there are few temporal
snapshots. However, none of these options is suitable to represent many
variables at the same time.

### 2018-08-15 paper/study.tex
missing the conclusions

\begin{figure*}
    \centering
 \includegraphics[width=0.85\linewidth]{newTeaser.pdf}
 \caption{Worflow to discover gentrification in Chicago: the purple cluster
 corresponds to high education / income. Its population is increasing over time,
 absorbing from the majority White cluster (orange). By selecting the purple
 cluster in 2010, the region is highlighted in the maps. The proportion of
 people with 4+ years of college is increasing in the whole city (grey IQRs),
 but significantly more in this region (black).\label{fig:chiWorkflow}}
\end{figure*}

@@

\subsection{Los Angeles}%TODO HERE
We selected a region around the metropolitan area of Los Angeles (LA), following
urban density, not administrative boundaries. The summary of the results using
all aspects with equal weights and eight clusters is illustrated in
Figure~\ref{fig:la}. The full statistical description of the clusters is
illustrated in Figure~\ref{fig:laAll}, where the most relevant aspect of each
cluster is highlighted. From the trajectories plot, we can see that there is a
large but shrinking cluster, depicted in green, one increasing cluster in
purple, an almost constant orange cluster, a smaller but increasing pink
cluster, and three other small clusters. The corresponding map illustrates where
these clusters are located, and that they are somewhat geographically stable,
with some movement between the green, orange, and purple clusters.

@@

From Figure~\ref{fig:laAll}, we can see that the green cluster is characterized by
a high percentage of White population, low percentage of population in the
lowest income bracket, mostly administrative occupations, and about 30\% of the
population with four or more years of college. The orange cluster is
characterized by a high percentage of Black population, with few people in the
highest range of income and education. The purple cluster corresponds to a high
concentration of "Other" in race, which includes Hispanic for this dataset, high
concentration of Laborers, and low education and income. The pink cluster
contains a high percentage of Asian population and about 30\% of the population
with four or more years of college. The light green cluster contains very few
people in the lower income bracket, mostly White population, with the highest
percentage of population with four or more years of college, working
administrative jobs, and a high concentration of singles. The yellow cluster
represents Black population, with higher level of education and income, mostly
working administrative jobs. The brown cluster represent a majority of single
population, working administrative jobs with mostly low income. The dark gray
cluster is characterized by all its population in the lowest income bracket, low
education level, with a majority of White population.

@@

Further, the larger clusters, green, orange, and purple, present a significant
intra-cluster variance in most variables, as indicated by the maximum and
minimum whiskers of the boxplots. For instance, while fifty percent of the CTs
in the green cluster have between 20\% and 40\% of people in the lowest income
bracket, that cluster also includes CTs where none and all the population
belongs to that bracket. This might indicate that this cluster could be further
divided, and currently represents different groups of people that are not
different enough to be separated into clusters in this level of the hierarchy.

While the larger clusters are expected, the light green, brown, and dark gray
clusters are more surprising. These clusters are small and ephemeral, including
only a few CTs, and are peculiarly different enough to be separated into their
own clusters, at this level of the hierarchy. While these small clusters might
be interesting enough to warrant further study, they might suppress larger, but
more subtle, clusters from appearing in this interface,  due to the limited
number of clusters.

@@

While Toronto is more dynamic than Los Angeles, possibly due to size
differences, the volatile regions shown in Figure~\ref{fig:toVol} did not change
as quickly or dramatically as the ones shown in Figure~\ref{fig:laVol}, which
involved twice as many people. We found this trend to be related to the
countries themselves, Canadian cities have larger areas undergoing slow, gradual
changes, whereas American cities have more general stability, but quicker
changes in smaller scales. The supplementary material contains brief summaries
of all the regions accessible in this prototype.

\section{Expert feedback}
%This could be a big step toward better understanding of evolving urban morphology.
To assess the proposed method, we contacted academic and industry experts in
sociology and urban sciences. The supplementary material contains all our
communication with five experts, identified from A to E.

Their response was positive,  mentioning that the prototype allows them to
analyze census data without the additional work of obtaining and cleaning the
data (A, B, E), and it allows the inclusion of geographic visual analysis tools
in their research process (D). It enables the users to tell different stories
about neighborhoods/cities and their changes (A), visualize the relationship
between key urban variables over time (D), offering a quick way to identify
particular neighborhoods that one may be interested in studying more in depth
around a particular issue or efficiently understanding the context of an area
(E).  Indeed, the experts identified gentrification processes in Manhattan (B)
and Dallas (E), reinforced a hypothesis for occupational clustering (D), and
highlighted how the method can be used to compare neighborhoods and cities (A).
These remarks attest that we satisfied the design requirements regarding the
existing demographic groups and their changes.

While the interface was "easy to navigate" (B), it was also considered
"overwhelming" (A), "intimidating" (E), and "tricky to interpret" (C), possible
side-effects of our effort to increase  representational accuracy, where we
avoided using simplified labels. Specifically, while identifying clusters by
their most relevant variables was welcome, the overlap of information from
different clusters in the boxplot was "a bit confusing" (C). While the map of
trajectories was mentioned as a "good summary map", how it related to the
clustering method was unclear (C). While we provide different options on how the
colors are used on this map, both are sub optimal, reliably representing several
distinct entities using colors is still an open problem in visualization.

The experts also mentioned the poor responsiveness of the method when changes in
the clustering parameters required server-side processing (B,D). Indeed, the
current implementation can take a few minutes to cluster regions with high
number of CTs, like Los Angeles or Brooklyn.

@@

More importantly, while these experts were aware that our method does not
perform geographic normalization, none of them mention it. We did not
specifically ask if this difference led to unexpected results, but rather
inquired if they found interesting insights, leading them to consider the
interface, its options, and results as a whole. Therefore the method was
neutrally compared against other, traditionally normalized results. We interpret
the fact that most of them were interested in the next steps as an affirmation
of the accuracy of the method.

>>>>

\begin{figure*}
    \centering
 \includegraphics[width=0.85\linewidth]{newTeaser.pdf}
 \caption{Workflow to discover gentrification in Chicago: the purple cluster
 corresponds to high education / income. Its population is increasing over time,
 absorbing from the majority White cluster (orange). By selecting the purple
 cluster in 2010, the region is highlighted in the maps. The proportion of
 people with 4+ years of college is increasing in the whole city (grey IQRs),
 but significantly more in this region (black).\label{fig:chiWorkflow}}
\end{figure*}

@@

\subsection{Los Angeles}
We selected a region around the metropolitan area of Los Angeles (LA), following
urban density. The summary of the results using all aspects with equal weights
and eight clusters is illustrated in Figure~\ref{fig:la}. The full statistical
description of the clusters is illustrated in Figure~\ref{fig:laAll}, where the
most relevant aspect of each cluster is highlighted. From the trajectories plot,
we can see that there is a large but shrinking cluster, depicted in green, one
increasing cluster in purple, an almost constant orange cluster, a smaller but
increasing pink cluster, and three other small clusters. The corresponding map
illustrates where these clusters are located, and that they are somewhat
geographically stable, with some movement between the green, orange, and purple
clusters.

@@

From Figure~\ref{fig:laAll}, we can see that the green cluster is characterized
by a high percentage of White population, low percentage of population in the
lowest income bracket, mostly administrative occupations, and about 30\% of the
population with four or more years of college. The orange cluster is
characterized by a high percentage of Black population, with few people in the
highest range of income and education. The purple cluster corresponds to a high
concentration of "Other" in race, which includes Hispanic for this dataset, high
concentration of Laborers, and low education and income. The pink cluster
contains a high percentage of Asian population and about 30\% of the population
with four or more years of college. The light green cluster contains very few
people in the lower income bracket, mostly White population, with the highest
percentage of population with four or more years of college, working
administrative jobs, and a high concentration of singles. The yellow cluster
represents Black population, with higher level of education and income, mostly
working administrative jobs. The brown cluster represent a majority of single
population, working administrative jobs with mostly low income. The dark gray
cluster is characterized by all its population in the lowest income bracket, low
education level, with a majority of White population. Since the extremes in the
boxplots of the grey cluster are not significantly different, we can also
surmise that this cluster is either small or homogeneous.

@@

The  green, orange, and purple clusters present a significant intra-cluster
variance in most variables, as indicated by extreme whiskers of the boxplots.
While fifty percent of the CTs in the green cluster have between 20\% and 40\%
of people in the lowest income bracket, that cluster also includes CTs where
none and all the population belongs to that bracket. This might indicate that
this cluster represents different groups of people that are not different enough
to be separated at this level of the hierarchy. Conversely, the light green,
brown, and dark gray clusters are different enough to be separated into their
own clusters at this level, despite being small and ephemeral, including only a
few CTs.

% While they might be interesting enough to
% warrant further study, they might suppress larger, but more subtle, clusters
% from appearing in this interface due to the limited number of clusters.

@@

\smallskip While Toronto is more dynamic than Los Angeles, possibly due to size
differences, the volatile regions shown in Figure~\ref{fig:toVol} did not change
as quickly or dramatically as the ones shown in Figure~\ref{fig:laVol}, which
involved twice as many people. We found this trend to be related to the
countries themselves, Canadian cities have larger areas undergoing slow, gradual
changes, whereas American cities have more general stability, but quicker
changes in smaller scales. The supplementary material contains brief summaries
of all the regions accessible in this prototype.

\section{Expert feedback}
%This could be a big step toward better understanding of evolving urban morphology.
To assess the proposed method, we contacted academic and industry experts in
sociology and urban sciences. The supplementary material contains the
communication with five experts, verbatim, identified from A to E.

Their overall response was positive,  mentioning that the prototype allows them
to analyze census data without the additional work of obtaining and cleaning the
data (A, B, E), and it allows the inclusion of geographic visual analysis tools
in their research process (D). It enables the users to tell different stories
about neighborhoods/cities and their changes (A), visualize the relationship
between key urban variables over time (D), offering a quick way to identify
particular neighborhoods that one may be interested in studying more in depth
around a particular issue or efficiently understanding the context of an area
(E).  Indeed, the experts identified gentrification processes in Manhattan (B)
and Dallas (E), reinforced a hypothesis for occupational clustering (D), and
highlighted how the method can be used to compare neighborhoods and cities (A).
In summary, the proposed methodology can be a viable alternative for the visual
analytics of evolving demographic data.

While the interface was "easy to navigate" (B), it was also considered
"overwhelming" (A), "intimidating" (E), and "tricky to interpret" (C), possible
side-effects of our effort to increase  representational accuracy, where we
avoided using simplified representation or labels. Identifying clusters by their
most relevant variables was welcome, but the overlap of information from
different clusters in the boxplot was "a bit confusing" (C) when color is not
present. Further, most clusters can be sufficiently characterized using only the
most relevant aspect, but this is not generally true.

While the map of trajectories was mentioned as a "good summary map", how it
related to the clustering method was unclear (C). While we provide different
options on how the colors are used, both are sub optimal since reliably
representing several distinct entities using colors is humanly unfeasible.
Indeed, the number of distinguishable colors was a significant constraint, we
found indications that more clusters should be used in some cases, even if eight
clusters is more than what is traditionally considered in these analyses.
Conversely, increasing the number of clusters would also complicate the
interpretation of the results.

The experts also mentioned the poor responsiveness of the method when changes in
the clustering parameters required server-side processing (B,D). Indeed, the
current implementation can take a few minutes to cluster regions with high
number of CTs, like Los Angeles or Brooklyn. Server-processing reduced the
amount of data transfered to client, but might increase response time under
load. We implemented a cache policy that greatly reduces the response time, but
fully pre-processing the results is not practical due to size of the parameter
space.

@@

More importantly, while these experts were aware that our method does not
perform geographic normalization, none of them mention it. We did not
specifically ask if this difference led to unexpected results, but rather
inquired if they found interesting insights, leading them to consider the
interface, its options, and results as a whole. Most experts found phenomena
corroborated by the specialized literature, indicating that our methodology
produces equivalent results, with a fraction of the effort. We interpret the
fact that most of them were interested in the next steps as an affirmation of
the accuracy of the method.

### 2018-08-15 sup/sup.tex
one past until LA
missing the conclusions

% We sent feedback requests to several researchers involved in Sociology and Urban
% sciences. Not all answered before the submission deadline, but we will keep an
% updated version of this document at:

% \href{http://ug.daniels.utoronto.ca/~diasf/CTEvo/sup.pdf}{\url{http://ug.daniels.utoronto.ca/~diasf/CTEvo/sup.pdf}}

% Last edit: \today

@@

\subsection{A - Cary Wu}
\href{https://soci.ubc.ca/persons/cary-wu/}{\url{https://soci.ubc.ca/persons/cary-wu/}}
% \subsection{Expert A}
\emph{Doctoral candidate in Sociology, UBC.}

@@

\subsection{B - Ethan A. Fosse}
% \href{https://scholar.harvard.edu/ethanfosse}{\url{https://scholar.harvard.edu/ethanfosse}}
% \subsection{Expert B}
\emph{Research Associate in  Sociology, Harvard}

@@

\subsection{C - Fernando Calderón Figueroa}
\emph{Sociology Department, University of Toronto}

@@

\subsection{E - James Murdoch}%, 
\emph{Policy Analyst, 2M Research}

>>>>

@@

\subsection{A - Cary Wu}
%\href{https://soci.ubc.ca/persons/cary-wu/}{\url{https://soci.ubc.ca/persons/cary-wu/}}
\emph{Doctoral candidate in Sociology at the University of British Columbia, Canada.}

@@

\subsection{B - Ethan A. Fosse}
% \href{https://scholar.harvard.edu/ethanfosse}{\url{https://scholar.harvard.edu/ethanfosse}}
\emph{Research Associate at Princeton University, USA}

@@

\subsection{C - Fernando Calderón Figueroa}
\emph{Doctoral candidate in Sociology at the University of Toronto, Canada}

@@

\subsection{E - James Murdoch}
\emph{Policy Analyst at 2M Research, USA}
